Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-06-07 paper/msg.tex
Complete first draft of MCDN description.

In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.  The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.

@@

However, we follow \cite{GraeserKornhuber2009} in decomposing a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.  That is, suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
\begin{equation}
\chi^J = b^J - s^J \in \mathcal{V}_0^J \label{eq:finedefectobstacle}
\end{equation}
be the associated defect obstacle, with $\chi^J \le 0$.  Note that $z^J \in \mathcal{V}_0^J$ is an admissible perturbation of $s^J$, i.e.~$s^J+z^J \in \mathcal{K}^J$, if and only if $z^J \ge \chi^J$.  Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$ \cite{GraeserKornhuber2009} on $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ by maximizing nodal values over the (open) supports of each each coarser-level hat function:
\begin{equation}
\mR z^j = \sum_{\ell=0}^{m_{j-1}-1} \max\left\{z^j[i] \,:\,\psi_\ell^{j-1}(x_i^j,y_i^j) \right\}\,\psi_\ell^{j-1}.  \label{eq:monotonerestriction}
\end{equation}
Now a defect obstacle is defined on each level:
\begin{equation}
\chi^{j-1} = \mR \chi^j  \label{eq:recursivedefectobstacle}
\end{equation}
for $j=1,\dots,J$, as shown for example in Figure \ref{fig:decompclassical}.  We also use the (nonpositive) differences,
\begin{equation}
\phi^j = \chi^j - \chi^{j-1},  \label{eq:downobstacles}
\end{equation}
with $\phi^0=\chi^0$.  Smoothers solve one-dimensional VIs on each level in different admissible sets as one goes down and up in the V-cycle; we define down- and up-constraint sets
\begin{equation}
\mathcal{D}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}, \qquad \mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \phi^j\right\}. \label{eq:downupsets}
\end{equation}

@@

FIXME MCD theory

FIXME DEBUG THIS PSEUDOCODE nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-vcycle}(J,s^J,b^J,\ell^J,\id{down}=0,\id{coarse}=1,\id{up}=1)\text{:} \\+
    $\chi^J = b^J - s^J$ \\
    $g^J = s^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \mR \chi^j$ \\
      $\phi^j = \chi^j - \chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smoother}}^{\text{\id{down}}}(y^j,g^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\
      $f = F^j(g^j+y^j) - \ell^j$ \\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = F^{j-1}(g^{j-1}) - R f$ \\-
    $y^0 = 0$ \\
    $\text{\pr{smoother}}^{\text{\id{coarse}}}(y^0,g^0,\ell^0,\chi^0)$ \qquad \ct{smoothing in $\mathcal{U}^0$} \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smoother}}^{\text{\id{up}}}(z^j,g^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{U}^j$} \\-
    return $z^J$
\end{pseudo*}
\caption{FIXME Nonlinear MCD V-cycle.}
\label{pc:mcdn-vcycle}
\end{pcode}

Finally, we need to be clear about when an iteration has converged.  For a discrete obstacle $\varphi^j \in \mathcal{V}^j$, an admissible current iterate $s^j\in \{r^j \ge \varphi^j\} \subset \mathcal{V}^j$, and a residual vector $F^j \in (\mathcal{V}^j)'$ let
\begin{equation}
\vertiii{F^j}_{(s^j,\varphi^j)} = \left(\sum_{s_i > \varphi_i} |F[\psi_i]|^2 + \sum_{s_i = \varphi_i} |\min\{F[\psi_i],0\}|^2\right)^{1/2} \label{eq:cpnorm}
\end{equation}
be the ``CP residual norm'' associated to the finite-dimensional complementarity problem (CP) $s^j \ge \varphi^j$, $F(s^j) \ge 0$, and $(s^j-\varphi^j) F(s^j) = 0$.  Note $s^j \in \mathcal{V}^j$ solves this CP if and only if $s^j$ is admissible and $\vertiii{F(s^j)}_{(s^j,\varphi^j)}=0$.  As shown in Pseudocode \ref{pc:mcdn-solver} we will iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-solver}(J,s^J,b^J,\id{atol}=10^{-20},\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
    function $\rho(s)[r] = F^J(s)[r] - \ip{a}{r}$ \qquad\qquad \ct{residual of \eqref{eq:fesigpweakform}} \\
    $\rho_0=\vertiii{\rho(s^J)}_{(s^J,b^J)}$ \\
    for $k=1,\dots,\id{cyclemax}$ \\+
        $\chi^J = b^J - s^J$ \qquad\qquad \ct{fine-level defect obstacle} \\
        $s^J \gets s^J + \pr{mcdn-vcycle}(J,s^J,\chi^J)$ \\
        $\rho_k=\vertiii{\rho(s^J)}_{(s^J,b^J)}$ \\
        if $\rho_k < \id{atol}$ or $\rho_k \le \id{rtol} \, \rho_0$ \\+
            break \\--
\end{pseudo*}
\caption{The SIGP is solved by iterating V-cycles (Pseudocode \ref{pc:mcdn-vcycle}) until the CP residual norm \eqref{eq:cpnorm} is small.}
\label{pc:mcdn-solver}
\end{pcode}

@@

\section*{Acknowledgments}  Thanks to David Maxwell for suggestions on the formulation and well-posedness of the model.

>>>>

In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.  The fundamental idea of such a scheme is the basic geometric multigrid idea that smoothing on coarse levels will rapidly reduce the high frequencies in the error, but here we must maintain admissibility on each level in a manner which does not re-introduce high frequencies.  Our method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.

@@

However, we follow \cite{GraeserKornhuber2009} in decomposing a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.  That is, suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
\begin{equation}
\chi^J = b^J - s^J \in \mathcal{V}_0^J \label{eq:finedefectobstacle}
\end{equation}
be the associated defect obstacle, with $\chi^J \le 0$.  Note that $z^J \in \mathcal{V}_0^J$ is an admissible perturbation of $s^J$, i.e.~$s^J+z^J \in \mathcal{K}^J$, if and only if $z^J \ge \chi^J$.  Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$ \cite{GraeserKornhuber2009} on $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ by maximizing nodal values over the (open) supports of each each coarser-level hat function:
\begin{equation}
\mR z^j = \sum_{\ell=0}^{m_{j-1}-1} \max\left\{z^j[i] \,:\,\psi_\ell^{j-1}(x_i^j,y_i^j) \right\}\,\psi_\ell^{j-1}.  \label{eq:monotonerestriction}
\end{equation}
Using $\mR$ a defect obstacle is defined on each level:
\begin{equation}
\chi^{j-1} = \mR \chi^j  \label{eq:recursivedefectobstacle}
\end{equation}
for $j=1,\dots,J$.  An example is shown in Figure \ref{fig:decompclassical}.  We also compute the differences,
\begin{equation}
\phi^j = \chi^j - \chi^{j-1},  \label{eq:downobstacles}
\end{equation}
with $\phi^0=\chi^0$ by definition; note $\phi^j\le 0$.

@@

A point-wise smoother on a given level will solve one-dimensional VIs at each node.  However, different admissible sets are used going down and up in the V-cycle.  Let
\begin{equation}
\mathcal{D}^j = \left\{y^j \in \mathcal{V}_0^j \,:\, y^j \ge \phi^j\right\}, \qquad \mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}. \label{eq:downupsets}
\end{equation}
The down-admissible constraint sets $\mathcal{D}^j$ sum to an up-admissible constraint set for each level, i.e.~$\mathcal{D}^0 + \mathcal{D}^1 + \dots + \mathcal{D}^j = \mathcal{U}^j$ in the sense that every $z^j \in \mathcal{U}^j$ is the sum of elements, though not uniquely.  The $\mathcal{U}^j$ are themselves nested in $\mathcal{U}^J= \{z^J \,:\, z^J \ge b^J - s^J\}$, and this forms a multilevel cone approximation, from the inside, of the continuum defect constraint set $\mathcal{K} - s = \{z\,:\,z \ge b - s\}$:
\begin{equation}
\mathcal{U}^0 \subset \mathcal{U}^1 \subset \dots \subset \mathcal{U}^J \subset \mathcal{K} - s \subset W_0^{1,\qq}(\Omega). \label{eq:innerconeapprox}
\end{equation}

We seek a solution to VI \eqref{eq:fesigpweakform} on the fine level.  The original idea of Tai \cite{Tai2003} is to use a multilevel constraint decomposition (MCD) like the above  system $\{\mathcal{D}^j\}$ to compute corrections $y^j \in \mathcal{D}^j$ which sum to a solution of the (nonlinear) fine-level VI,
\begin{equation}
F^J(s^J + y^J + \dots + y^j)[v^j - y^j] \ge (a,v^j - y^j) \quad \text{for all } v^j \in \mathcal{D}^j, \label{eq:mcdoriginal}
\end{equation}
where $F^J = F^h$ in \eqref{eq:fesigpweakform}.  In \cite{Tai2003}, VIs \eqref{eq:mcdoriginal} are to be solved on descending levels $j=J$ down to $j=0$, forming a down-slash or V(1,0) cycle.  (See also \cite[Algorithm 4.7]{GraeserKornhuber2009}.)  Note that $y^J + \dots + y^j \in \mathcal{U}^J$ so $s^J + y^J + \dots + y^j \ge b^J$ is admissible in the original sense, and that \eqref{eq:mcdoriginal} exploits the simplification $(s^J + y^J + \dots + v^j) - (s^J + y^J + \dots + y^j) = v^j - y^j$.

However, there is no practical way to achieve rapid solution of VI \eqref{eq:mcdoriginal} on coarse levels if the latest iterate $s^J + y^J + \dots + y^j$ must be represented; it has fine-level content.  This is the reason to add the full approximation scheme (FAS) \cite{Trottenbergetal2001} concept of a nonlinear coarse-level ``equation'', which is here a nonlinear VI.  In order to implement our approach we need to state the FAS coarse-level correction equation based on appropriate restrictions, as follows.

Let $\iR: \mathcal{V}^{j+1} \to \mathcal{V}^j$ be the injection restriction operator \cite{Trottenbergetal2001}.  Suppose $y^j \in \mathcal{D}^j$ denotes the desired down-correction on the $j$th level.  For corrections $y^j \in \mathcal{D}^j$, define $t^j \in \mathcal{V}^j$ as the $j$th-level approximation to the corrected fine-level solution (i.e.~$t^j \approx s^J + y^J + \dots + y^j$),
\begin{equation}
t^j = \begin{cases} s^J, & j=J \\
                    \iR(t^{j+1} + y^{j+1}), & j < J.
      \end{cases}  \label{eq:fassolution}
\end{equation}
Note that $t^j$ satisfies $t^j \ge \iR(\dots(\iR(b^J))\dots)$ by construction because $y^j \in \mathcal{D}^j$.  Though represented on a coarse level, $t^j$ has admissible point values relative to the fine bed $b^J$.

Let $F^j$ represent the FE discretization of $F$ (Section \ref{sec:fe}) on the $j$th level, and let $R: (\mathcal{V}^{j+1})^* \to (\mathcal{V}^j)^*$ be the canonical restriction on functionals \cite{GraeserKornhuber2009}.  The FAS coarse-level VI is
\begin{equation}
F^j(t^j+y^j)[v^j-y^j] \ge \ell^j[v^j-y^j] \quad \text{for all } v^j \in \mathcal{D}^j. \label{eq:fasequation}
\end{equation}
where
\begin{equation}
\ell^j[v] = \begin{cases} \ip{a}{v}, & j=J \\
                          F^j(t^j)[v] + R \left(\ell^{j+1} - F^{j+1}(t^{j+1}+y^{j+1})\right)[v], & j < J. \end{cases}  \label{eq:fasell}
\end{equation}
On the one hand, all quantities in VI \eqref{eq:fasequation} are represented on the $j$th level, without reference to finer-mesh data.  On the other hand, for $j<J$ the VI \eqref{eq:fasequation} can be rearranged using \eqref{eq:fasell} to a correction form like the FAS equation for PDEs (e.g.~\cite[equation (5.3.12)]{Trottenbergetal2001}),
\begin{equation}
F^j(t^j+y^j)[v^j-y^j] - F^j(t^j)[v^j-y^j] \ge R \left(\ell^{j+1} - F^{j+1}(t^{j+1}+y^{j+1})\right)[v^j-y^j], \label{eq:fasequationtraditional}
\end{equation}
for all $v^j \in \mathcal{D}^j$.  In \eqref{eq:fasequationtraditional} the functional on the left and right side should be smooth, thus explaining \eqref{eq:fasequation}.  Following the solution of \eqref{eq:fasequation} on all descending levels $j=J,\dots,0$, in fact done by a smoother on each level, a V(1,0) cycle is completed and the new fine-mesh iterate is $s^J + y^J + \dots + y^0$.

The above formulas suffice for a V(1,0) cycle, and represent the FAS extension of Algorithm 4.7 in \cite{GraeserKornhuber2009}, and of the method of Tai \cite{Tai2003}.  However, for better performance we note that on the ascending part of a V-cycle we can smooth in a less-constrained set, the larger set $\mathcal{U}^j \supset \mathcal{D}^j$.  Regarding admissibility, the difference between descending and ascending in a V-cycle is that in the former case there are yet-to-be-computed corrections which we must force into small-enough sets $\mathcal{D}^j$ so that their sum is still admissible.  When ascending we have already summed the coarser-level corrections so we can smooth the result in the larger, nested sets $\mathcal{U}^j$ without destroying the admissibility of the yet-finer corrections to come.

The above ideas define the following MCDN V-cycle Pseudocode \ref{pc:mcdn-vcycle}.  This algorithm calls a smoothers from Section \ref{sec:smoothers} and it uses the canonical prolongation $P:\mathcal{V}^j \to \mathcal{V}^{j+1}$.  The default settings correspond to a V(0,1) cycle which we have found to be most efficient in tests on the classical obstacle problem \cite{Bueler2022}.  The returned value is a correction to the current iterate $s^J$.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-vcycle}(J,s^J,b^J,\id{down}=0,\id{coarse}=1,\id{up}=1)\text{:} \\+
    $\chi^J, \,\ell^J, \,t^J = b^J - s^J, \,\ip{a}{\cdot}, \,s^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \mR \chi^j$ \\
      $\phi^j = \chi^j - \chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smoother}}^{\text{\id{down}}}(y^j,t^j,\ell^j,\phi^j)$ \qquad \qquad \ct{in $\mathcal{D}^j$} \\
      $t^{j-1} = \iR(t^j + y^j)$ \\
      $\ell^{j-1} = F^{j-1}(t^{j-1}) + R(\ell^j - F^j(t^j+y^j))$ \\-
    $y^0 = 0$ \\
    $\text{\pr{smoother}}^{\text{\id{coarse}}}(y^0,t^0,\ell^0,\chi^0)$ \qquad \qquad \ct{in $\mathcal{U}^0$} \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smoother}}^{\text{\id{up}}}(z^j,t^j,\ell^j,\chi^j)$ \qquad \qquad \ct{in $\mathcal{U}^j$} \\-
    return $z^J$
\end{pseudo*}
\caption{MCDN V-cycle.}
\label{pc:mcdn-vcycle}
\end{pcode}

To our knowledge two aspects of this MCDN V-cycle are new:
\begin{enumerate}
\item While Tai \cite{Tai2003} defines and proves the convergence of a nonlinear MCD method, the implementation of this method cannot achieve $O(m_J)$ time for each V-cycle in nonlinear cases because the method refers to the fine level for the residual evaluation on each coarser level.  Our FAS-type modification achieve $O(m_J)$ time per V-cycle when applied to a local nonlinear problem, for example in the SIA model application in \cite{Bueler2022} in which the smoother is $O(m_j)$ on each level.  In our current Stokes case the smoother is nonlocal and $O(m_j)$ smoother time is not achieved.
\item Our V($\alpha$,$\beta$) cycles, with $\alpha$ down-smoother and $\beta$ up-smoother applications, are more efficient when $\beta >0$ than the method proposed in \cite{GraeserKornhuber2009} for $V(1,1)$ cycles because the smoothing occurs in a less-constrained set.
\end{enumerate}

Finally we consider when iterated V-cycles have converged.  For any discrete obstacle $\chi^j \in \mathcal{V}^j$, admissible iterate $z^j\in \{r^j \ge \chi^j\} \subset \mathcal{V}^j$, and residual vector $r^j \in (\mathcal{V}^j)'$ let
\begin{equation}
\vertiii{r^j}_{(z^j,\chi^j)} = \left(\sum_{z_i > \chi_i} |r^j[\psi_i^j]|^2 + \sum_{z_i = \chi_i} |\min\{r^j[\psi_i^j],0\}|^2\right)^{1/2}, \label{eq:cpnorm}
\end{equation}
where $\psi_i^j$ denotes a $j$th-level nodal basis function.  This defines a ``CP residual norm'' associated to the finite-dimensional complementarity problem $z^j \ge \chi^j$, $r^j(z^j) \ge 0$, and $(z^j-\chi^j) r^j(z^j) = 0$.  Note $z^j \in \mathcal{V}^j$ solves this CP if and only if $z^j$ is admissible and $\vertiii{r^j(z^j)}_{(z^j,\chi^j)}=0$.

As shown in Pseudocode \ref{pc:mcdn-solver}, we iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-solver}(J,s^J,b^J,\id{atol}=10^{-20},\id{rtol}=10^{-3},\id{cyclemax}=100)\text{:} \\+
    $\rho_0=\vertiii{(a,\cdot) - F^J(s^J)[\cdot]}_{(s^J,b^J)}$ \\
    for $k=1,\dots,\id{cyclemax}$ \\+
        $s^J \gets s^J + \pr{mcdn-vcycle}(J,s^J,b^J)$ \\
        $\rho_k=\vertiii{(a,\cdot) - F^J(s^J)[\cdot]}_{(s^J,b^J)}$ \\
        if $\rho_k < \id{atol}$ or $\rho_k \le \id{rtol} \, \rho_0$ \\+
            break \\--
\end{pseudo*}
\caption{The SIGP is solved by iterating V-cycles (Pseudocode \ref{pc:mcdn-vcycle}) until the CP residual norm \eqref{eq:cpnorm} is small.}
\label{pc:mcdn-solver}
\end{pcode}

@@

\section*{Acknowledgments}  Thanks to David Maxwell for discussions regarding the formulation and well-posedness of the model.
